Northern Pikeminnow Predation on Juvenile Chinook Salmon and Potential Impacts to Local Populations

### 2021-11-19 analysis/paper/paper.Rmd
added citations and re-knit .docx

An estimated 200 million juvenile salmonids emigrate through the lower Snake and Columbia rivers, annually. Of those, approximately 16.4 million (8%) are consumed by northern pikeminnow [Beamesderfer1996]. @Peterson1994 estimated that the annual loss of juvenile salmonids to northern pikeminnow to be 7.3% of all juvenile salmonids entering the John Day Reservoir. Additionally, 78% of these salmonids consumed near a dam were consumed when alive [@Peterson1994]. Dams are known to be areas of predation, but @Ward1995 estimated that 48% of predation is occurring away from dams in mid-reservoir areas. Most of the research has been targeted around dams and reservoirs, but there are areas functioning similar to a reservoir in the rivers upstream. One of these areas is the Deadwater Slough in the Salmon River. The Deadwater Slough is a 25.3-acre area that is unnaturally slow and deep within the Salmon River and resembles a small reservoir (Figure \@ref(fig:map)). This unnatural area has little to no cover, higher depth, and fine substrate which favor piscivorous fish predators (e.g., northern pikeminnow *Ptychocheilus oregonensis*, smallmouth bass *Micropterus dolomieu*) [@Watkins2015]. @Axel2015 demonstrated decreased rates of emigration and apparent survival for Sockeye Salmon *Oncorhynchus nerka* from Redfish Lake emigrating during the spring. Further, recent winter telemetry studies have indicated decreased transition probabilities (approximately 10% less than surrounding reaches) of juvenile Chinook salmon through Deadwater Slough during fall and winter months [@Ackerman2018; @Porter2019].

@@

To estimate the impacts from northern pikeminnow on juvenile Chinook Salmon outmigrants in Deadwater Slough, we used an R-based application of Fish Bioenergetics v4.0 developed by @Deslauriers2017. We estimated a daily rate of consumption for an individual northern pikeminnow based on predator and prey energy densities, water temperatures, and predator start and end weights. We also ran two alternative models one with a year of temperature inputs from 2019, and another with the same temperature inputs but a 10% growth in weight of the average pikeminnow at the end of the model. The average length of northern pikeminnow captured at the Deadwater Slough during our study was 370.6 mm which calculates to an average starting weight of 847 g using the FSA package (@Ogle2019) in R which provides a weight-length formula for northern pikeminnow sourced from @Parker1995. Predator energy density for northern pikeminnow was available in the Bioenergetics v4.0 application and was fixed at 6,703 Joules(J)/g. Prey energy densities was taken from @Moss2016 where they estimated juvenile Chinook Salmon at 21,500 J/g. Using these temperatures, weights, and energy inputs, we estimated a total consumption in grams for an individual norther pikeminnow within Deadwater Slough.

>>>>

An estimated 200 million juvenile salmonids emigrate through the lower Snake and Columbia rivers, annually. Of those, approximately 16.4 million (8%) are consumed by northern pikeminnow [Beamesderfer1996]. @Petersen1994 estimated that the annual loss of juvenile salmonids to northern pikeminnow to be 7.3% of all juvenile salmonids entering the John Day Reservoir. Additionally, 78% of these salmonids consumed near a dam were consumed when alive [@Petersen1994]. Dams are known to be areas of predation, but @Ward1995 estimated that 48% of predation is occurring away from dams in mid-reservoir areas. Most of the research has been targeted around dams and reservoirs, but there are areas functioning similar to a reservoir in the rivers upstream. One of these areas is the Deadwater Slough in the Salmon River. The Deadwater Slough is a 25.3-acre area that is unnaturally slow and deep within the Salmon River and resembles a small reservoir (Figure \@ref(fig:map)). This unnatural area has little to no cover, higher depth, and fine substrate which favor piscivorous fish predators (e.g., northern pikeminnow *Ptychocheilus oregonensis*, smallmouth bass *Micropterus dolomieu*) [@Watkins2015]. @Axel2015 demonstrated decreased rates of emigration and apparent survival for Sockeye Salmon *Oncorhynchus nerka* from Redfish Lake emigrating during the spring. Further, recent winter telemetry studies have indicated decreased transition probabilities (approximately 10% less than surrounding reaches) of juvenile Chinook salmon through Deadwater Slough during fall and winter months [@Ackerman2018; @Porter2019].

@@

To estimate the impacts from northern pikeminnow on juvenile Chinook Salmon outmigrants in Deadwater Slough, we used an R-based application of Fish Bioenergetics v4.0 developed by @Deslauriers2017. We estimated a daily rate of consumption for an individual northern pikeminnow based on predator and prey energy densities, water temperatures, and predator start and end weights. We also ran two alternative models one with a year of temperature inputs from 2019, and another with the same temperature inputs but a 10% growth in weight of the average pikeminnow at the end of the model. The average length of northern pikeminnow captured at the Deadwater Slough during our study was 370.6 mm which calculates to an average starting weight of 847 g using the FSA package (@Ogle2021) in R which provides a weight-length formula for northern pikeminnow sourced from @Parker1995. Predator energy density for northern pikeminnow was available in the Bioenergetics v4.0 application and was fixed at 6,703 Joules(J)/g. Prey energy densities was taken from @Moss2016 where they estimated juvenile Chinook Salmon at 21,500 J/g. Using these temperatures, weights, and energy inputs, we estimated a total consumption in grams for an individual norther pikeminnow within Deadwater Slough.
